Classless Java: Tuning Java Interfaces

### 2016-03-17 OOPSLA16/sections/Introduction.tex
Polished introduction

Object oriented languages always strive to offer great code reuse,
they aim to couple flexibility and rigour, expressive power and modular reasoning.
Two main idea have emerged to this end: prototype based (PB) and class based (CB) languages.
In prototype based languages an objects inherits from other objects,
and thus objects own both behaviour and state (and objects is all you have).

In class based languages an object is instance of a specific class, and classes inherits from other classes.
Here objects own the state, while classes contains behaviour and the structure of the state.

We present here a third alternative: the concept of
 \textbf{Interface based object oriented languages} (IB), where 
  objects implements interfaces, and they own implementation for the behaviour,
that is structurally defined in their interface.

  The literature provide good examples on how easy and modular can be combining
  multiple sources of pure behaviour (traits[]), and how
  hard it is to modularly combine multiple sources of behaviour \textbf{and} state (multiple inheritance[]).

  In IB, state is just a special kind of behaviour implementation.
  Object are the only responsible to define 
  the ultimate behaviour of a method and, for example, if such method is just a setter.
  Anything related to state is completely contained in the instances and do not leak in the interfaces.

@@

We propose a programming style, where we never use classes\footnote{More precisely, we never use the \Q@class@ keyword.}.
We call this restricted version of Java \emph{Class-less Java}.
Our proposed style enables powerful object-oriented idioms, similar to 
(stateful) multiple inheritance and field type refinement 
(allowing a kind of covariant setters refinement).

As for CB, in IB programmers do not define objects directly,but delegate the task to
\emph{object interfaces}, whose role is similar to non abstract classes in CB.

The idea is that all abstract operations in an Object interface will be
interpreted as \emph{abstract state operations};  The abstract state
operations include various common utility methods (such as getters and
setters, or clone-like methods). In the presence of subtyping, such
operations often require special care, as their types need to be
refined. Object interfaces provide support for type-refinement and can
automatically produce code that deals with type-refinement
adequately, while Java (and Scala) require explicit type-refinement.

@@

The Java annotation processors allows us to stick to pure Java8: by
annotating with \Q|@Obj| certain interfaces
we can generate boring and repetitive code
allowing interface instantiation and type refinement; such code should not be needed in the first place
in a real IB language.
%Since Java was not designed to be used in this way, our style can be verbose, especially about object instantiation.

@@

We formally define the behaviour of our \Q|Obj| annotation and we provide a
prototype
implementation using Java annotations to do AST
rewriting, allowing existing Java tools (such as IDEs) to work
out-of-the-box with our implementation. As a result, we could
experiment object interfaces with several interesting Java programs,
and conduct various case studies.

@@

To evaluate the usefulness of object interfaces, we illustrate 3
applications. The first application is a simple 
solution to the Expression Problem~\cite{wadler98expression}, supporting independent 
extensibility~\cite{zenger05independentlyextensible}, and without boilerplate code. The second
application shows how embedded DSLs using fluent interfaces~\cite{fowler2005fluentinterface} 
can be easily defined using object interfaces. The last
application is a larger case study for a simple Maze game implemented with 
multiple inheritance. For the last application we show that there is a
significant reduction in the numbers of lines of code when compared 
to an existing implementation~\cite{bono14} using plain Java 8. 
Noteworthy, all applications are implemented 
without defining a single class!

@@

\item{\bf Type preservation guarantees:}
We discuss our formalization of a subset of Java8 type system and how we use this
to characterize safety properties about our annotations.
%\item {\bf Classless Java (CJ):} A simple formal calculus that models 
%the essential features of Java 8 interfaces with default methods, and 
%can be used to formally define the translation of object interfaces. 
%We prove several properties of the translation\footnote{Proofs and prototype implementation are available in
%  the supplementary materials.}.

>>>>

Object oriented languages have always strived to offer great code reuse.
They aim to couple flexibility and rigour, expressive power and
modular reasoning.  Two main ideas have emerged to this end: prototype
based (PB)~\cite{} and class based (CB) languages~\cite{}.  In prototype based
languages objects inherit from other objects, and thus objects own
both behaviour and state (and objects is all you have).
In class based languages an object is instance of a specific class,
and classes inherit from other classes.  Here objects own the state,
while classes contain behaviour and the structure of the state.

We present here a third alternative: the concept of \textbf{Interface
  based object oriented languages} (IB), where objects implements
interfaces, and they own the implementation for the behaviour, which
is structurally defined in their interface.

The literature provide good examples on how easy and modular can be
combining multiple sources of pure behaviour (traits~\cite{}), and how hard
it is to modularly combine multiple sources of behaviour \textbf{and}
state (multiple inheritance~\cite{}).\bruno{needs strenghtning}

To retain the easiness and modularity of combining multiple sources of
pure behaviour, in IB state is just a special kind of behaviour
implementation. Objects are the only responsibles to define the
ultimate behaviour of a method and, for example, if such method is
just a setter. Anything related to state is completely contained in
the instances and do not leak in the interfaces.\bruno{Not sure if
  this is totally clear. Implicit here is the fact that objects are
  never defined explicitly, I think. And the behaviour is introduced 
implicitly.}

@@

In the context of Java, what we propose is essentially a programming
style, where we never use classes\footnote{More precisely, we never
  use the \Q@class@ keyword.}.  We call this restricted version of
Java \emph{Class-less Java}.  Our proposed style enables powerful
object-oriented idioms, similar to (stateful) multiple inheritance and
field type refinement (allowing a kind of covariant setters
refinement).

As for CB, in IB programmers do not define objects directly, but
delegate the task to \emph{object interfaces}, whose role is similar
to non abstract classes in CB.

To model state, the idea is that all abstract operations in an Object
interface will be interpreted as \emph{abstract state operations}; The
abstract state operations include various common utility methods (such
as getters and setters, or clone-like methods). In the presence of
subtyping, such operations often require special care, as their types
need to be refined. Object interfaces provide support for
type-refinement and can automatically produce code that deals with
type-refinement adequately, while Java (and Scala) require explicit
type-refinement.

@@

Using Java annotation processors, we produce an implementation of
Classless Java, which allow us to stick to pure Java8. By annotating
with \Q|@Obj| the interfaces that represent object interfaces, we can
generate boring and repetitive code for interface instantiation and
type refinement.  Such code should not be needed in the first place in
a real IB language.

%Since Java was not designed to be used in this way, our style can be verbose, especially about object instantiation.

@@

We formally define the behaviour of our \Q|Obj| annotation and we provide a
prototype implementation using Java annotations to do AST
rewriting, allowing existing Java tools (such as IDEs) to work
out-of-the-box with our implementation. As a result, we
experiment object interfaces with several interesting Java programs,
and conduct various case studies.

@@

To evaluate the usefulness of object interfaces, we illustrate
3\bruno{needs updates}
applications. The first application is a simple 
solution to the Expression Problem~\cite{wadler98expression}, supporting independent 
extensibility~\cite{zenger05independentlyextensible}, and without boilerplate code. The second
application shows how embedded DSLs using fluent interfaces~\cite{fowler2005fluentinterface} 
can be easily defined using object interfaces. The last
application is a larger case study for a simple Maze game implemented with 
multiple inheritance. For the last application we show that there is a
significant reduction in the numbers of lines of code when compared 
to an existing implementation~\cite{bono14} using plain Java 8. 
Noteworthy, all applications are implemented 
without defining a single class!

@@

\item{\bf Type preservation guarantees:}\bruno{needs rephrasing}
We discuss our formalization of a subset of Java8 type system and how we use this
to characterize safety properties about our annotations.
%\item {\bf Classless Java (CJ):} A simple formal calculus that models 
%the essential features of Java 8 interfaces with default methods, and 
%can be used to formally define the translation of object interfaces. 
%We prove several properties of the translation\footnote{Proofs and prototype implementation are available in
%  the supplementary materials.}.
